Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-22 paper.tex

For this study, we applied the bio2zarr compression to the chromosome 20 exome-like 
data from the All of Us dataset. The exome-like data is constructed by extracting from all WGS samples
the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
transcripts. This dataset, stored as a single compressed VCF file of 7.44 GiB, 
includes 715,256 variants, of which more than 700 are classified as pathogenic or likely pathogenic.

The compression was performed on the All of Us Researcher Workbench platform 
(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB 
of RAM, which cost \$3.80/hour. We first converted the VCF to an ICF, which 
took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU 
time, achieving a parallel efficiency of 86\%. This step incurred a cost of 
\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed 
across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with 
a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a 
cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across 
232,809 chunk files.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced 
    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-top-fields-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
% /call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
% /call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
% /call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
% /call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
% /call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
% /call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
% /call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
\bottomrule
\end{tabular}
\end{table}

@@

Large genomes from non-model organisms pose particular problems for
analysis workflows that often are tailored for human conditions. For
instance, the binary Sequence Alignment/Map index format BAI has a 512
Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
The spruce BAM files used for variant calling were BAI indexed, which
motivated the partitioning into multiple VCF files in the first place,
adding complexity to downstream processing.

A similar limitation is imposed by the VCF POS field which is encoded
in 32-bit format and can hold chromomosomes up to 2.14 Gbp. Although
spruce does not have chromosomes that exceed the limit, other species
do, such as the recently sequence mistletoe, with chromosomes larger
than 8 Gbp~\cite{dtol_mistletoe}. Zarr does not impose this
limitation, and the coordinate system could easily be converted to
true genome coordinates for storage in Zarr. This updating procedure
could be applied to any species of choice.

>>>>

For this study, we converted the chromosome 20 exome-like 
data from the All of Us dataset to VCF Zarr.
The exome-like data is constructed by extracting from all WGS samples
the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
transcripts. This dataset, stored as a single compressed VCF file of 7.44 GiB, 
includes 715,256 variants, of which more than 700 are classified as pathogenic or likely pathogenic.

Conversion was performed on the All of Us Researcher Workbench platform 
(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB 
of RAM, which cost \$3.80/hour. We first converted the VCF to  ICF, which 
took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU 
time, achieving a parallel efficiency of 86\%. This step incurred a cost of 
\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed 
across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with 
a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a 
cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across 
232,809 chunk files.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns 
    for the All of Us exome-like data. The chunk size is
    1,000 variants $\times$ 10,000 samples. 
    For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-auo-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
/call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
/call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
/call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
/call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
/call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
/call\_FT & object & 498.61 MiB & 2700.0 & 6.00\% \\
\bottomrule
\end{tabular}
\end{table}

[TODO discuss local alleles and mention Table~\ref{tab-auo-data}].

@@

Large genomes from non-model organisms pose particular problems for
analysis workflows that often are tailored for human conditions. For
instance, the binary Sequence Alignment/Map index format
BAI~\cite{li2009sequence} has a 512
Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
The spruce BAM files used for variant calling were BAI indexed, which
motivated the partitioning into multiple VCF files in the first place,
adding complexity to downstream processing.

A similar limitation is imposed by the VCF POS field which is encoded
in 32-bit format and can hold chromomosomes up to 2.14 Gbp. Although
spruce does not have chromosomes that exceed the limit, other species
do, such as the recently sequence mistletoe, with chromosomes larger
than 8 Gbp~\cite{dtol_mistletoe}. Zarr's flexible type 
system means that coordinates can be stored as 64 or 32 bit integers
as needed, and there is therefore no particular limit on genome size.
% limitation, and the coordinate system could easily be converted to
% true genome coordinates for storage in Zarr. This updating procedure
% could be applied to any species of choice.
